---
title: "Far-Right in Scandinavia"
subtitle: "The Far Right (Week 4)"
author: "Dr. Christos Vrakopoulos"
format:
  revealjs:
    theme: solarized
    slide-number: true
    incremental: true
    transition: slide
    chalkboard: true    
    fontsize: 26px
    css: custom.css
editor: visual
---

```{r setup, include=FALSE}
# Load required packages
library(dplyr)
library(ggplot2)
library(plotly)
library(tidyverse)
library(manifestoR)
library(lubridate)
library(ggthemes)

# Set Manifesto Project API key
mp_setapikey(key = "193c54d64b1189c526214074350ead07")

# Load Manifesto Project data
mp_data <- manifestoR::mp_maindataset()

# Define party codes for Scandinavian far-right parties
# Sweden: Sweden Democrats (SD)
# Denmark: Danish People's Party (DPP), Progress Party
# Norway: Progress Party (FrP)
# Finland: Finns Party (PS/True Finns)

party_codes <- tribble(
  ~party, ~party_label,
  # Sweden
  11710, "SD",        # Sweden Democrats (was wrong: 11820)
  # Denmark  
  13720, "DF",        # Danish People's Party (correct)
  13951, "FrP-DK",    # Progress Party (Denmark) (was wrong: 13620)
  # Norway
  12951, "FrP-NO",    # Progress Party (Norway) / Anders Lange's Party (was wrong: 12720)
  # Finland
  14820, "PS"         # True Finns / Finns Party (correct)
)

# Extract far-right party data
far_right_raw <- mp_data %>%
  filter(party %in% party_codes$party) %>%
  mutate(year = year(as.Date(edate))) %>%
  left_join(party_codes, by = "party") %>%
  select(countryname, year, edate, party_label, pervote, absseat) %>%
  filter(!is.na(pervote)) %>%
  arrange(countryname, year, edate)

# Combine vote shares within same election
far_right_combined <- far_right_raw %>%
  group_by(countryname, year, edate) %>%
  summarise(
    vote_share = sum(pervote, na.rm = TRUE),
    parties = paste(unique(party_label), collapse = " + "),
    .groups = "drop"
  ) %>%
  group_by(countryname, year) %>%
  slice_max(edate, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(country = countryname, year, vote_share)

far_right_data <- far_right_combined

# Individual country datasets
sweden_data <- far_right_raw %>% 
  filter(countryname == "Sweden") %>%
  group_by(countryname, year, edate) %>%
  summarise(vote_share = sum(pervote, na.rm = TRUE), .groups = "drop") %>%
  group_by(countryname, year) %>%
  slice_max(edate, n = 1) %>%
  ungroup() %>%
  select(country = countryname, year, vote_share)

denmark_data <- far_right_raw %>% 
  filter(countryname == "Denmark") %>%
  group_by(countryname, year, edate) %>%
  summarise(vote_share = sum(pervote, na.rm = TRUE), .groups = "drop") %>%
  group_by(countryname, year) %>%
  slice_max(edate, n = 1) %>%
  ungroup() %>%
  select(country = countryname, year, vote_share)

norway_data <- far_right_raw %>% 
  filter(countryname == "Norway") %>%
  group_by(year, edate) %>%
  summarise(vote_share = sum(pervote, na.rm = TRUE), .groups = "drop") %>%
  group_by(year) %>%
  slice_max(edate, n = 1) %>%
  ungroup() %>%
  mutate(country = "Norway") %>%
  select(country, year, vote_share)

finland_data <- far_right_raw %>% 
  filter(countryname == "Finland") %>%
  group_by(year, edate) %>%
  summarise(vote_share = sum(pervote, na.rm = TRUE), .groups = "drop") %>%
  group_by(year) %>%
  slice_max(edate, n = 1) %>%
  ungroup() %>%
  mutate(country = "Finland") %>%
  select(country, year, vote_share)
```

## Today's Lecture

::: incremental
-   **Part 1:** Why Scandinavia?
-   **Part 2:** Four Trajectories
-   **Part 3:** Comparative Synthesis
-   **Part 4:** Implications
:::

# Part 1: Why Scandinavia?


## The Scandinavian Puzzle

::::::: columns
:::: {.column width="50%"}
::: {.fragment .fade-in}
**Should be resistant to far-right:**
:::

::: incremental
-   Generous welfare states

-   Low inequality

-   High trust societies

-   Strong social democratic tradition

-   Consensus politics
:::
::::

:::: {.column width="50%"}

::: {.fragment .fade-in}
**Yet far-right succeeded:**
:::

::: incremental
-   Sweden: 20%+ (2018-)

-   Denmark: 21% peak (2015)

-   Norway: 22% peak (2009)

-   Finland: 19% (2011, 2019)
:::
::::
:::::::

# Part 2: Four Trajectories

## Sweden: Sweden Democrats (Sverigedemokraterna) {background-image="images/sd_logos.png" background-size="cover" background-opacity="0.2"}

::::: columns
::: {.column width="50%"}

::: {.fragment .fade-in}
#### From Fringe to Mainstream:
:::

-   Founded 1988, roots in neo‑Nazi skinhead milieu

-   Early years: explicit racist and white‑supremacist rhetoric

-   Visible street‑movement style: uniforms, marches, episodes of violence

-   1995-2005: Marginalized (\<1%)

::: {.fragment .fade-in}
#### Transformation (2005-):
:::

-   Jimmie Åkesson becomes leader (2005)

-   Professionalization strategy (message discipline, media training, local branches)

-   Expel extremists

-   Moderate image, suit-and-tie

-   Goal: become a ‘normal’ party voters can support without stigma
:::

::: {.column width="50%"}

::: {.fragment .fade-in}
#### What They Stand For Today:
:::

-   **Nativism:** Immigration framed as threat to “Swedish values”

-   **Economic:** Welfare chauvinism. “Strong welfare, but for Swedes first”

-   **Social:** Socially conservative

-   **EU:** Soft Eurosceptic (critical of EU integration, not exit)

::: {.fragment .fade-in}
**Key insight:** Longest organizational transformation in Europe (20+ years from neo-Nazi to 20% party)
:::
:::
:::::

## Sweden: Trajectory & Pattern

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

p_sweden <- ggplot(sweden_data, aes(x = year, y = vote_share)) +
  geom_line(linewidth = 1.5, color = "#FDB913") +
  geom_point(size = 3, color = "#FDB913") +
  geom_vline(xintercept = c(2005, 2010), linetype = "dashed", alpha = 0.5) +
  annotate("text", x = 2005, y = 2, label = "Åkesson\nleader", hjust = -0.1, size = 4) +
  annotate("text", x = 2010, y = 7, label = "Breakthrough", hjust = -0.1, size = 4) +
  annotate("text", x = 2022, y = 21, label = "20.5%", hjust = -0.1, size = 4) +
  labs(x = "Year", y = "Vote Share (%)") +
  theme_solarized(base_size = 16)

ggplotly(p_sweden, tooltip = c("x", "y")) %>%
  layout(autosize = FALSE, width = 900, height = 500)
```

## Denmark: Danish People's Party (Dansk Folkeparti) {background-image="images/df.png" background-size="cover" background-opacity="0.2"}

::::: columns
::: {.column width="50%"}
::: {.fragment .fade-in}
#### Origins:
:::

-   Founded 1995 (split from Progress Party)

-   Pia Kjærsgaard leader (1995-2012)

-   Moderate from the start (learning from Norwegian FrP’s radicalism)

::: {.fragment .fade-in}
#### Breakthrough:
:::

-   1998: 7.4%

-   2001: 12% support center-right govt

-   2015: **21.1%** (peak, second largest)
:::

::: {.column width="50%"}

::: {.fragment .fade-in}
#### Core Ideology:
:::

-   **Nativism:** Danish culture under threat

-   **Economic:** Radical welfare chauvinism (often cited as Europe’s strongest)

-   **Social:** Traditionalist, Christian-inflected social conservatism

-   **EU:** Eurosceptic

::: {.fragment .fade-in}
#### 2019: COLLAPSE
:::
::: incremental
-   Social Democrats adopt DPP’s immigration/welfare positions

-   DPP collapses: 21% → 8.7% (2015–2019)

-   Mainstream accommodation works in Denmark

:::
:::
:::::

## Denmark: Trajectory & Pattern

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

p_denmark <- ggplot(denmark_data, aes(x = year, y = vote_share)) +
  geom_line(linewidth = 1.5, color = "#C60C30") +
  geom_point(size = 3, color = "#C60C30") +
  annotate("rect", xmin = 2015, xmax = 2019, ymin = 0, ymax = 25, alpha = 0.1, fill = "blue") +
  annotate("text", x = 2017, y = 17, label = "Social Democrats\nsteal positions", 
           size = 3.5, color = "blue") +
  geom_vline(xintercept = 2015, linetype = "dashed", alpha = 0.5) +
  annotate("text", x = 2015, y = 22, label = "Peak\n21.1%", hjust = -0.1, size = 4, fontface = "bold") +
  labs(x = "Year", y = "Vote Share (%)") +
  theme_solarized(base_size = 16)

ggplotly(p_denmark, tooltip = c("x", "y")) %>%
  layout(autosize = FALSE, width = 900, height = 500)
```

## Norway: Progress Party (Fremskrittspartiet) {background-image="images/npp.png" background-size="cover" background-opacity="0.2"}

:::::: columns
::: {.column width="50%"}

::: {.fragment .fade-in}
#### Origins:
:::

-   Founded 1973 (anti-tax, libertarian)

-   Anders Lange: "Strong reduction in taxes"

-   1980s: Added anti-immigration (critique of refugee/asylum policy)

-   Long presence (50+ years)

-   Gradual rise: from protest party to 2nd largest in 2009

::: {.fragment .fade-in}
#### Government (2013-2020):
:::

-   Coalition with Conservatives

-   First far-right in Scandinavian govt

-   Moderated tone, but kept restrictive immigration stance
:::

:::: {.column width="50%"}

::: {.fragment .fade-in}
#### Core Ideology:
:::

-   **Nativism:** Immigration control

-   **Economic:** Neoliberal + welfare chauvinist; Protect oil wealth for ‘Norwegians first’

-   **Social:** Mixed (liberal on some issues)

-   **EU:** Norway outside EU (Eurosceptic on EEA)

::: {.fragment .fade-in}
**Result:** Moderation in government → loss of protest appeal → electoral decline → rise again
:::

-   2009: 22.9% (peak)

-   2021: 11.6% (after government)

-   2025: 23.8%
::::
::::::

## Norway: Trajectory & Pattern

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

p_norway <- ggplot(norway_data, aes(x = year, y = vote_share)) +
  geom_line(linewidth = 1.5, color = "#BA0C2F") +
  geom_point(size = 3, color = "#BA0C2F") +
  annotate("rect", xmin = 2013, xmax = 2020, ymin = 0, ymax = 25, alpha = 0.15, fill = "red") +
  annotate("text", x = 2016.5, y = 23, label = "In\ngovernment", size = 3.5, color = "red") +
  geom_vline(xintercept = 2009, linetype = "dashed", alpha = 0.5) +
  annotate("text", x = 2009, y = 24, label = "Peak\n22.9%", hjust = -0.1, size = 4, fontface = "bold") +
  labs(x = "Year", y = "Vote Share (%)") +
  theme_solarized(base_size = 16)

ggplotly(p_norway, tooltip = c("x", "y")) %>%
  layout(autosize = FALSE, width = 900, height = 500)
```

## Finland: Finns Party (Perussuomalaiset) {background-image="images/Perussuomalaiset_Logo.svg.png" background-size="cover" background-opacity="0.2"}

::::::: columns
:::: {.column width="50%"}

::: {.fragment .fade-in}
#### Origins:
:::

-   Founded 1995 (Finnish Rural Party successor)

-   Agrarian populism + nationalism

-   Timo Soini leader (1997-2017)

::: {.fragment .fade-in}
#### Breakthrough:
:::

-   2011: **19.1%** (EU bailout crisis)

-   2015: **17.7%** - enters government

::: {.fragment .fade-in}
**2017: THE SPLIT**
:::

-   Cost of office: coalition compromises vs. radical base expectations

-   Party splits: Blue Reform (moderates, stayed in govt) vs. Finns Party (radicalized opposition)
::::

:::: {.column width="50%"}

::: {.fragment .fade-in}
#### Core Ideology:
:::

-   **Nativism:** Finnish culture, anti-immigration

-   **Economic:** Welfare chauvinist, Eurosceptic

-   **Social:** Socially conservative

-   **EU:** Hard Eurosceptic (anti-bailout)

::: {.fragment .fade-in}
**After split:**
:::

-   Blue Reform: Moderate, stayed in govt → died

-   Finns Party: Radical, opposition → organizationally purified, electoral recovery (2023: 20.1%, second largest party)
::::
:::::::

## Finland: Trajectory & Pattern

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

p_finland <- ggplot(finland_data, aes(x = year, y = vote_share)) +
  geom_line(linewidth = 1.5, color = "#003580") +
  geom_point(size = 3, color = "#003580") +
  geom_vline(xintercept = c(2011, 2017), linetype = "dashed", alpha = 0.5) +
  annotate("text", x = 2011, y = 20, label = "Breakthrough\n19.1%", hjust = -0.1, size = 4, fontface = "bold") +
  annotate("text", x = 2017, y = 15, label = "Split", hjust = -0.1, size = 4) +
  annotate("rect", xmin = 2015, xmax = 2017, ymin = 0, ymax = 20, alpha = 0.1, fill = "red") +
  annotate("text", x = 2016, y = 18, label = "In govt", size = 3, color = "red") +
  labs(x = "Year", y = "Vote Share (%)") +
  theme_solarized(base_size = 16)

ggplotly(p_finland, tooltip = c("x", "y")) %>%
  layout(autosize = FALSE, width = 900, height = 500)
```

# Part 3: Comparative Synthesis

## Scandinavian Patterns

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

# Add flag emojis
far_right_flags <- far_right_data %>% 
  filter(country %in% c("Sweden", "Denmark", "Norway", "Finland")) %>%
  mutate(
    flag_emoji = case_when(
      country == "Sweden" ~ "🇸🇪",
      country == "Denmark" ~ "🇩🇰",
      country == "Norway" ~ "🇳🇴",
      country == "Finland" ~ "🇫🇮"
    )
  )

# Get only the last year for each country
last_points <- far_right_flags %>%
  group_by(country) %>%
  slice_max(year, n = 1) %>%
  ungroup()

p <- ggplot(far_right_flags, 
            aes(x = year, y = vote_share, color = country, group = country)) +
  geom_line(linewidth = 1.5, alpha = 0.8) +
  geom_point(size = 3, alpha = 0.7) +
  geom_text(data = last_points,
            aes(label = flag_emoji), 
            size = 8, 
            hjust = -0.5) +
  scale_color_manual(
    values = c("Sweden" = "#FDB913", "Denmark" = "#C60C30", 
               "Norway" = "#BA0C2F", "Finland" = "#003580")
  ) +
  labs(x = "Year", y = "Vote Share (%)", color = "Country") +
  theme_solarized(base_size = 16) +
  theme(legend.position = "bottom", legend.title = element_text(face = "bold"))

ggplotly(p, tooltip = c("x", "y", "colour")) %>%
  layout(hovermode = "x unified", autosize = FALSE, width = 900, height = 500)
```

## Welfare Chauvinism in Action

| Country | Welfare State Type | Far-Right Strategy | Outcome |
|------------------|------------------|------------------|------------------|
| Sweden | Universal, generous | "Swedish welfare for Swedes" | 20%+ sustained |
| Denmark | Universal, generous | Strictest in Europe | 21% peak, then theft |
| Norway | Universal, generous | Oil fund + immigration | 23% peak, govt decline |
| Finland | Universal, generous | Anti-EU bailout + immigration | 19% breakthrough |

::: {.fragment .fade-in}
Generous welfare states don't prevent far-right - they provide **justification for exclusion**
:::

## Government Participation Effects

| Country | In Government?      | Duration | Result                 | Vote Change     |
|--------------|---------------|--------------|-----------------|--------------|
| Sweden  | Support only (2024) | Ongoing  | TBD                    | \-              |
|         |                     |          |                        |                 |
| Denmark | Never               | \-       | Position theft instead | 21% → 2.6%      |
| Norway  | Yes (2013-2020)     | 7 years  | Gradual decline        | 22.9% → 11.6%   |
| Finland | Yes (2015-2017)     | 2 years  | Split, moderates died  | 17.7% → (split) |

::: {.fragment .fade-in}
**Pattern:** Government participation **harmful** for far-right across all cases
:::

::: {.fragment .fade-in}
**Exception:** Can organizational strength cushion the blow? (Norway vs. Austria)
:::

## Who Votes Far-Right in Scandinavia?

```{r}
#| echo: false
#| warning: false
#| fig-width: 12
#| fig-height: 8

load("/Users/cvrakopo/My Drive/Publications/ESS Data 1-10/Final Creation/ess_analysis_full.RData")  # Adjust path as needed

# Filter data for the five countries
ess_scan <- ess_analysis %>%
  filter(country %in% c("Finland", "Denmark", "Sweden", "Norway"))


# Prepare data with age categories and calculate differences from reference
demo_comparison <- ess_scan %>%
  mutate(
    # Create age categories
    age_cat = case_when(
      agea < 35 ~ "18-34",
      agea >= 35 & agea < 55 ~ "35-54",
      agea >= 55 & agea < 70 ~ "55-69",
      agea >= 70 ~ "70+"
    ),
    
    # Immigration attitudes
    immi_cat = case_when(
      is.na(p_immi_index) ~ NA_character_,
      p_immi_index < 30 ~ "Liberal",
      p_immi_index >= 30 & p_immi_index < 70 ~ "Neutral",
      p_immi_index >= 70 ~ "Restrictive"
    ),
    
    # Trust in politicians (0-10 scale)
    trust_cat = case_when(
      is.na(trstplt) ~ NA_character_,
      trstplt <= 3 ~ "Low trust",
      trstplt >= 4 & trstplt <= 6 ~ "Medium trust",
      trstplt >= 7 ~ "High trust"
    ),
    
    # Left-Right scale (0-10 scale)
    lr_cat = case_when(
      is.na(lrscale) ~ NA_character_,
      lrscale <= 3 ~ "Left",
      lrscale >= 4 & lrscale <= 6 ~ "Center",
      lrscale >= 7 ~ "Right"
    ),
    
    # Household income feelings (1-4 scale)
    income_cat = case_when(
      is.na(hincfel) ~ NA_character_,
      hincfel == 1 ~ "Comfortable",
      hincfel == 2 ~ "Coping",
      hincfel == 3 ~ "Difficult",
      hincfel == 4 ~ "Very difficult"
    ),
    
    # Simplify education
    education_simple = case_when(
      educ_level %in% c("Less than lower secondary", "Lower secondary") ~ "Low",
      educ_level %in% c("Lower tier upper secondary", "Upper tier upper secondary", 
                        "Advanced vocational, sub-degree") ~ "Medium",
      educ_level %in% c("Lower tertiary education, BA level", 
                        "Higher tertiary education, >= MA level") ~ "High"
    ),
    
    # Simplify class
    class_simple = case_when(
      oesch8 %in% c("Production workers", "Service workers") ~ "Working class",
      oesch8 %in% c("Clerks", "Small business owners") ~ "Lower middle",
      oesch8 %in% c("Technical (semi-)professionals", "(Associate) managers") ~ "Upper middle",
      oesch8 %in% c("Socio-cultural (semi-)professionals", 
                    "Self-employed professionals and large employers") ~ "Professional/Managers"
    )
  ) %>%
  # Convert far_right to numeric if it's not already
  mutate(far_right_num = as.numeric(far_right == "Far-right"))

# Calculate proportions for each category
calc_props <- function(data, var) {
  data %>%
    filter(!is.na(.data[[var]])) %>%  # Remove NAs for this variable
    group_by(across(all_of(var))) %>%
    summarise(
      prop = mean(far_right_num, na.rm = TRUE) * 100,
      n = n(),
      se = sd(far_right_num, na.rm = TRUE) / sqrt(n()) * 100,
      .groups = "drop"
    ) %>%
    rename(category = 1)
}

# Get proportions for each demographic
gender_props <- calc_props(demo_comparison, "gender") %>% mutate(variable = "Gender")
age_props <- calc_props(demo_comparison, "age_cat") %>% mutate(variable = "Age")
immi_props <- calc_props(demo_comparison, "immi_cat") %>% mutate(variable = "Immigration attitudes")
trust_props <- calc_props(demo_comparison, "trust_cat") %>% mutate(variable = "Political trust")
lr_props <- calc_props(demo_comparison, "lr_cat") %>% mutate(variable = "Left-Right position")
income_props <- calc_props(demo_comparison, "income_cat") %>% mutate(variable = "Economic situation")
edu_props <- calc_props(demo_comparison, "education_simple") %>% mutate(variable = "Education")
urban_props <- calc_props(demo_comparison, "urban_rural") %>% mutate(variable = "Location")
class_props <- calc_props(demo_comparison, "class_simple") %>% mutate(variable = "Class")

# Combine all
all_props <- bind_rows(
  gender_props, age_props, immi_props, trust_props, lr_props, income_props,
  edu_props, urban_props, class_props
) %>%
  filter(!is.na(category))

# Set reference categories and calculate differences
reference_levels <- tribble(
  ~variable, ~reference,
  "Gender", "Female",
  "Age", "18-34",
  "Immigration attitudes", "Liberal",
  "Political trust", "High trust",
  "Left-Right position", "Center",
  "Economic situation", "Comfortable",
  "Education", "High",
  "Location", "Urban",
  "Class", "Professional/Managers"
)

# Get reference values
reference_values <- all_props %>%
  inner_join(reference_levels, by = "variable") %>%
  filter(category == reference) %>%
  select(variable, ref_prop = prop)

# Calculate differences from reference
plot_data <- all_props %>%
  left_join(reference_values, by = "variable") %>%
  mutate(
    diff = prop - ref_prop,
    ci_lower = diff - 1.96 * se,
    ci_upper = diff + 1.96 * se,
    is_reference = category %in% reference_levels$reference
  ) %>%
  # Order categories within each variable
  mutate(
    category = factor(category, levels = c(
      # Gender
      "Male", "Female",
      # Age
      "70+", "55-69", "35-54", "18-34",
      # Immigration
      "Restrictive", "Neutral", "Liberal",
      # Trust
      "Low trust", "Medium trust", "High trust",
      # Left-Right
      "Right", "Center", "Left",
      # Economic situation
      "Very difficult", "Difficult", "Coping", "Comfortable",
      # Education
      "Low", "Medium", "High",
      # Location
      "Rural", "Urban",
      # Class
      "Working class", "Lower middle", "Upper middle", "Professional/Managers"
    )),
    variable = factor(variable, levels = c(
      "Gender", "Age", "Immigration attitudes", "Political trust", 
      "Left-Right position", "Economic situation", "Education", "Location", "Class"
    ))
  ) %>%
  arrange(variable, category)

# Create the plot
ggplot(plot_data, aes(x = diff, y = category)) +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray40", linewidth = 0.8) +
  geom_segment(aes(x = 0, xend = diff, y = category, yend = category, 
                   color = diff > 0, alpha = !is_reference), 
               linewidth = 4) +
  geom_errorbar(aes(xmin = ci_lower, xmax = ci_upper, alpha = !is_reference), 
                orientation = "y", width = 0.2, linewidth = 0.8) +
  geom_point(aes(alpha = !is_reference), size = 2.5, color = "black") +
  geom_text(data = plot_data %>% filter(is_reference),
            aes(label = "Reference"), x = 0.5, size = 3, hjust = 0, 
            fontface = "italic", color = "gray50") +
  facet_wrap(~variable, ncol = 1, scales = "free_y", strip.position = "left") +
  scale_color_manual(values = c("TRUE" = "#E74C3C", "FALSE" = "#3498DB"), guide = "none") +
  scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.3), guide = "none") +
  scale_x_continuous(
    breaks = seq(-10, 15, 5),
    labels = function(x) paste0(ifelse(x > 0, "+", ""), x, " pp")
  ) +
  labs(
    title = "Far-Right Support: Differences from Reference Categories",
    subtitle = "Denmark, Finland, Norway, and Sweden (ESS data)",
    x = "← Less likely to support far-right          More likely to support far-right →",
    y = NULL
  ) +
  theme_solarized(base_size = 14) +
  theme(
    strip.placement = "outside",
    strip.text.y.left = element_text(face = "bold", angle = 0, hjust = 1, size = 13),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.x = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "gray40", size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(margin = margin(t = 10), size = 12)
  )

```

## Key Puzzles from Scandinavia

::::::::::: columns
:::::: {.column width="50%"}
::: {.fragment .fade-in}
1.  The Professionalization Puzzle\*\*
:::

-   Can far-right parties successfully transform from extremist to mainstream?

::: {.fragment .fade-in}
2.  The Welfare Paradox
:::

-   Do generous universal welfare states contain or fuel far-right support?

::: {.fragment .fade-in}
3.  The Government Curse
:::

-   Does entering government harm far-right parties and under what conditions??
::::::

:::::: {.column width="50%"}
::: {.fragment .fade-in}
4.  The Position Theft Puzzle
:::

-   Under what conditions mainstream parties eliminate far-right by adopting their positions?

::: {.fragment .fade-in}
5.  The Moderation Trap
:::

-   Does moderation in office weaken far-right parties—or make them more acceptable?

::: {.fragment .fade-in}
6.  The Permanence Question
:::

-   Are far-right parties crisis-driven protest vehicles, or durable party-system players?
::::::
:::::::::::

## Next Week: Southern Europe

**The crisis puzzle:**

::: incremental
-   Greece: Severe crisis → Golden Dawn surge
-   Spain: Severe crisis → No far-right until 2018 (10-year delay)
-   Portugal: Crisis → Far-right decline initially
-   Italy: Different pattern (post-fascist trajectory)
:::

. . .

**Questions:** Why did economic crisis produce such different outcomes?

**Seminar:** Country team presentations + discuss organizational development and government participation patterns

## Questions?
