---
title: "Far-Right in Southern Europe"
subtitle: "The Far Right (Week 5)"
author: "Dr. Christos Vrakopoulos"
format:
  revealjs:
    theme: solarized
    slide-number: true
    incremental: true
    
    transition: fade
    fontsize: 26px
    css: custom.css
editor: visual
---

```{r setup, include=FALSE}
# Load required packages
library(dplyr)
library(ggplot2)
library(plotly)
library(tidyverse)
library(manifestoR)
library(lubridate)
library(ggthemes)

# Set Manifesto Project API key
mp_setapikey(key = "193c54d64b1189c526214074350ead07")

# Load Manifesto Project data
mp_data <- manifestoR::mp_maindataset()

# Define party codes for Southern European far-right parties
party_codes <- tribble(
  ~party, ~party_label,
  # Italy
  32710, "MSI/AN",    # Italian Social Movement → National Alliance (same code)
  32720, "Lega",      # Northern League/Lega
  32630, "FdI",       # Brothers of Italy
  # Spain
  33710, "VOX",       # Vox
  # Greece
  34710, "LAOS",      # Popular Orthodox Rally
  34720, "GD",        # Golden Dawn (Chrysi Avgi)
  34730, "ANEL",      # Independent Greeks
  34711, "EL",        # Greek Solution (Elliniki Lysi)
  # Portugal
  35720, "Chega",     # Enough
  35951, "PSN"        # National Solidarity Party
)

# Extract far-right party data
far_right_raw <- mp_data %>%
  filter(party %in% party_codes$party) %>%
  mutate(year = year(as.Date(edate))) %>%
  left_join(party_codes, by = "party") %>%
  select(countryname, year, edate, party_label, pervote, absseat) %>%
  filter(!is.na(pervote)) %>%
  arrange(countryname, year, edate)

# Combine vote shares within same election
far_right_combined <- far_right_raw %>%
  group_by(countryname, year, edate) %>%
  summarise(
    vote_share = sum(pervote, na.rm = TRUE),
    parties = paste(unique(party_label), collapse = " + "),
    .groups = "drop"
  ) %>%
  group_by(countryname, year) %>%
  slice_max(edate, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(country = countryname, year, vote_share)

far_right_data <- far_right_combined

# Individual country datasets
italy_data <- far_right_raw %>% 
  filter(countryname == "Italy") %>%
  group_by(countryname, year, edate) %>%
  summarise(vote_share = sum(pervote, na.rm = TRUE), .groups = "drop") %>%
  group_by(countryname, year) %>%
  slice_max(edate, n = 1) %>%
  ungroup() %>%
  select(country = countryname, year, vote_share)

spain_data <- far_right_raw %>% 
  filter(countryname == "Spain") %>%
  group_by(countryname, year, edate) %>%
  summarise(vote_share = sum(pervote, na.rm = TRUE), .groups = "drop") %>%
  group_by(countryname, year) %>%
  slice_max(edate, n = 1) %>%
  ungroup() %>%
  select(country = countryname, year, vote_share)

greece_data <- far_right_raw %>% 
  filter(countryname == "Greece") %>%
  group_by(year, edate) %>%
  summarise(vote_share = sum(pervote, na.rm = TRUE), .groups = "drop") %>%
  group_by(year) %>%
  slice_max(edate, n = 1) %>%
  ungroup() %>%
  mutate(country = "Greece") %>%
  select(country, year, vote_share)

portugal_data <- far_right_raw %>% 
  filter(countryname == "Portugal") %>%
  group_by(year, edate) %>%
  summarise(vote_share = sum(pervote, na.rm = TRUE), .groups = "drop") %>%
  group_by(year) %>%
  slice_max(edate, n = 1) %>%
  ungroup() %>%
  mutate(country = "Portugal") %>%
  select(country, year, vote_share)
```

## Today's Lecture

::: incremental
-   **Part 1:** The Crisis Puzzle
-   **Part 2:** Four Trajectories
-   **Part 3:** Comparative Synthesis
-   **Part 4:** Implications
:::

# Part 1: The Crisis Puzzle

## Why Southern Europe?

::::::: columns
:::: {.column width="50%"}

::: {.fragment .fade-in}
#### Shared context:
:::

-   Fascist/authoritarian past (Mussolini, Franco, Salazar, Greek junta)

-   Late democratization (1970s)

-   **2008-2015: Severe economic crisis**

-   High unemployment, austerity, EU bailouts
::::

:::: {.column width="50%"}
::: {.fragment .fade-in}
#### Yet outcomes vary:
:::

::: incremental
-   **Greece**: "Early" and violent breakthrough – Golden Dawn \~7% (2012–15)

-   **Spain**: Delayed entry – Vox only after 2018

-   **Italy**: Party mutation – Lega + rise of FdI, now governing

-   **Portugal**: Late breakthrough – Chega after 2019, from marginal to third largest party
:::
:::: 
::::::: 

# Part 2: Four Trajectories

## Italy: Post-Fascist Trajectory {background-image="images/meloni.jpg" background-size="cover" background-opacity="0.2"}

::::: columns
::: {.column width="50%"}

::: {.fragment .fade-in}
#### Three waves:
:::

1.  **MSI (1946-1995)**: Post-fascist nostalgia
    -   Founded by Mussolini supporters
    -   Anti-communist, anti-democratic
    -   5-8% stable, South-based
2.  **AN (1995-2009)**: Legitimization
    -   Fini's moderation: "post-fascist"
    -   Joined Berlusconi governments → Normalisation
    -   Merged into PdL (2009)
:::

::: {.column width="50%"}
3.  **Lega (1991-)**: From regional protest to national radical right (but not post‑fascist)
    -   Bossi (1991-2012): Northern separatism
    -   Salvini (2013-): National, anti-immigration
    -   Rebrand: Lega Nord → Lega
4.  **FdI (2012-)**: Neo-fascist heir
    -   Meloni: Ex-MSI youth wing
    -   2022: **26%** - Prime Minister
    -   **First far-right PM since Mussolini**
:::
:::::

## Italy: Trajectory & Pattern

```{r}
#| echo: false
#| warning: false
#| fig-width: 12
#| fig-height: 8

p_italy <- ggplot(italy_data, aes(x = year, y = vote_share)) +
  geom_line(linewidth = 1.5, color = "#009246") +
  geom_point(size = 3, color = "#009246") +
  geom_vline(xintercept = c(1995, 2013, 2022), linetype = "dashed", alpha = 0.5) +
  annotate("text", x = 1995, y = 18, label = "AN forms", hjust = -0.1, size = 4) +
  annotate("text", x = 2013, y = 12, label = "Salvini\nleader", hjust = -0.1, size = 4) +
  annotate("text", x = 2022, y = 28, label = "Meloni\nPM", hjust = -0.1, size = 4, fontface = "bold") +
  labs(x = "Year", y = "Vote Share (%)") +
  theme_solarized(base_size = 16)

ggplotly(p_italy, tooltip = c("x", "y")) %>%
  layout(autosize = FALSE, width = 900, height = 500)
```

## Spain: Forty Years Without a Major Far Right Party {background-image="images/Vox.jpg" background-size="cover" background-opacity="0.2"}

::::: columns
::: {.column width="50%"}
::: {.fragment .fade-in}
**Franco legacy (1939-1975):**
:::
-   Falangist movement (openly fascist)

-   36-year dictatorship

-   Strong Church–state alliance, anti-communist, nationalist


::: {.fragment .fade-in}
**Transition (1975-1978):**
::: 
-   Regime members joined mainstream

-   **People's Party (PP)** incorporated Francoists

-   Far-right electorally absorbed by PP
:::

::: {.column width="50%"}

::: {.fragment .fade-in}
**Failed parties (1975-2013):**
:::
-   Multiple neo-fascist groups: \<1%

-   Stigma too strong (Franco memory)

::: {.fragment .fade-in}
**Vox (2013-):**
:::
-   Founded by ex-PP members (Abascal)

-   2013-2017: \<1%

-   **2018: 10.3%** -\> Breakthrough

-   **Catalan crisis as catalyst**: nationalism + territorial unity
:::
:::::

-   Why a territorial crisis (Catalonia) helps the far-right?

## Spain: Trajectory & Pattern

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# Note: Spain data will be very sparse (Vox only 2018+)
p_spain <- ggplot(spain_data, aes(x = year, y = vote_share)) +
  geom_line(linewidth = 1.5, color = "#AA151B") +
  geom_point(size = 3, color = "#AA151B") +
  geom_vline(xintercept = 2008, linetype = "dashed", color = "darkgray", alpha = 0.5) +
  geom_vline(xintercept = 2017, linetype = "dashed", color = "black") +
  annotate("text", x = 2008, y = 5, label = "Crisis\n(no far-right)", hjust = -0.1, size = 3.5, color = "darkgray") +
  annotate("text", x = 2017, y = 8, label = "Catalan\ncrisis", hjust = -0.1, size = 4, fontface = "bold") +
  annotate("text", x = 2018, y = 12, label = "Vox\nbreakthrough", hjust = -0.1, size = 4) +
  labs(x = "Year", y = "Vote Share (%)") +
  theme_solarized(base_size = 16)

ggplotly(p_spain, tooltip = c("x", "y"))%>%
  layout(autosize = FALSE, width = 900, height = 500)
```

## Greece: Crisis-Driven Surge {background-image="images/gd.jpg" background-size="cover" background-opacity="0.2"}

::::: columns
::: {.column width="50%"}

::: {.fragment .fade-in}
**Military junta (1967-1974):**
:::
-   Papadopoulos regime: "Greece for Christian Greeks"

-   Collapsed after Cyprus invasion (1974)



::: {.fragment .fade-in}
**Failed parties (1974-2012):**
:::
-   National Alignment (1977): Merged with ND

-   Political Spring (1993): Macedonia issue

-   LAOS (2000-2012): Moderate far-right, 5-6%, part of coalition government with PASOK and ND

-   Independent Greeks (2012): Eurosceptic, governed with SYRIZA


:::

::: {.column width="50%"}

::: {.fragment .fade-in}
**Golden Dawn (1985-):**
:::
-   Founded by Michaloliakos (neo-Nazi)

-   2010–2012: Austerity, unemployment, anti‑bailout anger → protest vote

-   **Openly neo-Nazi:** Uniforms, salutes, violence

-   2013: Murder of Pavlos Fyssas → criminal organization trial

-   Leaders imprisoned, party collapsed

-   **Greek Solution (2019-) and other smaller parties**: Parliamentary, softer style; no open Nazi symbolism


:::
:::::

## Greece: Trajectory & Pattern

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

p_greece <- ggplot(greece_data, aes(x = year, y = vote_share)) +
  geom_line(linewidth = 1.5, color = "#0D5EAF") +
  geom_point(size = 3, color = "#0D5EAF") +
  geom_vline(xintercept = c(2010, 2013), linetype = "dashed", alpha = 0.5) +
  annotate("rect", xmin = 2010, xmax = 2015, ymin = 0, ymax = 35, alpha = 0.1, fill = "red") +
  annotate("text", x = 2012.5, y = 33, label = "Crisis peak\n25% unemployment", size = 3.5, color = "darkred") +
  annotate("text", x = 2012, y = 25, label = "GD surge\n7%", hjust = -0.1, size = 4, fontface = "bold") +
  annotate("text", x = 2013, y = 18, label = "Murder→\nInvestigation", hjust = -0.1, size = 3.5) +
  annotate("text", x = 2019, y = 7, label = "GD collapse\nGreek Solution", hjust = 1.1, size = 3.5) +
  labs(x = "Year", y = "Vote Share (%)") +
  theme_solarized(base_size = 16)

ggplotly(p_greece, tooltip = c("x", "y"))%>%
  layout(autosize = FALSE, width = 900, height = 500)
```

## Portugal: Why did Portugal resist far-right breakthrough for four decades? {background-image="images/chega.png" background-size="cover" background-opacity="0.2"}

::::: columns
::: {.column width="50%"}

::: {.fragment .fade-in}
**Salazar dictatorship (1932-1974):**
:::
-   Estado Novo: Authoritarian, Catholic, pro-Franco

-   Suppressed explicitly fascist rivals; regime monopolized authoritarian right



::: {.fragment .fade-in}
**Carnation Revolution (1974):**
:::
-   **Left-leaning** military coup

-   Peaceful transition

-   De-colonization

-   Left-leaning political culture (strong PCP, PS dominance, weak post-authoritarian right)

:::

::: {.column width="50%"}
::: {.fragment .fade-in}
**Far-right marginalized (1974-2019):** No successful radical right in parliament until 2019
:::
-   National Renovator Party (2000-): Extreme-right, \<1%


::: {.fragment .fade-in}
**Chega (2019-):**
:::
-   Founded by André Ventura (ex-PSD)

-   2019: 1.3%

-   2022: **7.2%** -\> Breakthrough

-   2024: **18%** -\> Third largest

-   **Finally broke through** (45 years after democratization)

:::
:::::

## Portugal: Trajectory & Pattern

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# Note: Portugal data will be very sparse (Chega only 2019+)
p_portugal <- ggplot(portugal_data, aes(x = year, y = vote_share)) +
  geom_line(linewidth = 1.5, color = "#FF0000") +
  geom_point(size = 3, color = "#FF0000") +
  geom_hline(yintercept = 0, color = "gray", linewidth = 0.5) +
  geom_vline(xintercept = 2011, linetype = "dashed", color = "darkgray", alpha = 0.5) +
  annotate("text", x = 2011, y = 3, label = "Crisis\n(no far-right)", hjust = -0.1, size = 3.5, color = "darkgray") +
  annotate("text", x = 2019, y = 2, label = "Chega\nfounded", hjust = -0.1, size = 4) +
  annotate("text", x = 2022, y = 8, label = "Breakthrough\n7.2%", hjust = -0.1, size = 4, fontface = "bold") +
  annotate("text", x = 2005, y = 12, label = "45 years: NO far-right\n(1974-2019)", hjust = 0.5, size = 4, fontface = "italic", color = "darkgray") +
  labs(x = "Year", y = "Vote Share (%)") +
  ylim(0, 15) +
  theme_solarized(base_size = 16)

ggplotly(p_portugal, tooltip = c("x", "y"))%>%
  layout(autosize = FALSE, width = 900, height = 500)
```

# Part 3: Comparative Synthesis

## The Southern European Crisis Puzzle

| Country      | Crisis Severity       | Far-Right 2010-2012 | Far-Right 2018-2024 |
|-----------------|--------------------|------------------|------------------|
| **Greece**   | Severe (bailout)      | 7% (GD surge)       | 4% (after collapse) |
| **Spain**    | Severe (bailout)      | 0% (no party)       | 15% (Vox)           |
| **Portugal** | Severe (bailout)      | 0% (no party)       | 18% (Chega)         |
| **Italy**    | Moderate (stagnation) | 5-15% (gradual)     | 35% (Lega+FdI)      |

. . .

**The puzzle:** Same economic crisis (2008-2015), yet dramatically different far-right trajectories. What explains this variation?

## South European Patterns

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 5

# Add flag emojis
far_right_flags <- far_right_data %>% 
  filter(country %in% c("Italy", "Spain", "Greece", "Portugal")) %>%
  mutate(
    flag_emoji = case_when(
      country == "Italy" ~ "🇮🇹",
      country == "Spain" ~ "🇪🇸",
      country == "Greece" ~ "🇬🇷",
      country == "Portugal" ~ "🇵🇹"
    )
  )

# Get only the last year for each country
last_points <- far_right_flags %>%
  group_by(country) %>%
  slice_max(year, n = 1) %>%
  ungroup()

p <- ggplot(far_right_flags, 
            aes(x = year, y = vote_share, color = country, group = country)) +
  geom_line(linewidth = 1.5, alpha = 0.8) +
  geom_point(size = 3, alpha = 0.7) +
  geom_vline(xintercept = 2008, linetype = "dashed", color = "red", linewidth = 1) +
  annotate("text", x = 2008, y = 25, label = "Financial\nCrisis", hjust = -0.1, color = "red", size = 4) +
  geom_text(data = last_points,
            aes(label = flag_emoji), 
            size = 8, 
            hjust = -0.5) +
  scale_color_manual(
    values = c("Italy" = "#009246", "Spain" = "#AA151B", 
               "Greece" = "#0D5EAF", "Portugal" = "#FF0000")
  ) +
  labs(x = "Year", y = "Vote Share (%)", color = "Country") +
  theme_solarized(base_size = 16) +
  theme(legend.position = "bottom", legend.title = element_text(face = "bold"))

ggplotly(p, tooltip = c("x", "y", "colour")) %>%
  layout(hovermode = "x unified", autosize = FALSE, width = 900, height = 500)
```

## Historical Legacies Matter

::::::: columns
:::: {.column width="50%"}
**How regimes ended:**

::: incremental
-   **Italy**: Mussolini defeated (war), gradual acceptance
-   **Spain**: Franco died, transition incorporated regime members
-   **Greece**: Junta collapsed (Cyprus), some integration
-   **Portugal**: Left-wing revolution, anti-fascist consensus
:::
::::

:::: {.column width="50%"}
**Effect on far-right:**

::: incremental
-   **Italy**: Longest continuous legitimization (MSI→AN→FdI)
-   **Spain**: Absorption into PP, blocked independent far-right for \~40 years
-   **Greece**: Multiple attempts, crisis breakthrough
-   **Portugal**: Strongest post-authoritarian stigma, 45-year delay
:::
::::
:::::::

. . .

**Key insight:** Different transitions create different party-space for the far right!!

**BUT**: How and why the effect of legacies break?

## Mainstream Responses Varied

| Country | Mainstream Strategy | Far-Right Result | Effectiveness |
|------------------|------------------|------------------|------------------|
| **Italy** | Accommodation (Berlusconi + AN) | Legitimized, now in government | Failed (legitimized) |
| **Spain** | Incorporation (Franco → PP) | Blocked 40 years, then breakthrough | Worked until 2018 |
| **Greece** | Partial integration (junta → ND) | Multiple attempts, crisis surge | Partially worked |
| **Portugal** | Isolation + left consensus | Blocked 45 years, then breakthrough | Worked until 2019 |

## Recent Convergence (2019-2024)

**All four now have far-right in parliament:**

::: incremental
-   **Italy**: Brothers of Italy (26%) - **Prime Minister Meloni**
-   **Spain**: Vox (12-15%) - Third largest, regional governments
-   **Greece**: Fragmentation
-   **Portugal**: Chega (18%) - Third largest, rapid rise
:::

. . .

**Pattern:** Historical resistance **delayed** but didn't **prevent** far-right emergence

. . .

**2010s:** Convergence across Europe - far-right now mainstream everywhere

## Who Votes Far-Right in Southern Europe?

```{r}
#| echo: false
#| warning: false
#| fig-width: 12
#| fig-height: 8

load("/Users/cvrakopo/My Drive/Publications/ESS Data 1-10/Final Creation/ess_analysis_full.RData")  # Adjust path as needed

# Filter data for the five countries
ess_south <- ess_analysis %>%
  filter(country %in% c("Greece", "Italy", "Spain", "Portugal"))


# Prepare data with age categories and calculate differences from reference
demo_comparison <- ess_south %>%
  mutate(
    # Create age categories
    age_cat = case_when(
      agea < 35 ~ "18-34",
      agea >= 35 & agea < 55 ~ "35-54",
      agea >= 55 & agea < 70 ~ "55-69",
      agea >= 70 ~ "70+"
    ),
    
    # Immigration attitudes
    immi_cat = case_when(
      is.na(p_immi_index) ~ NA_character_,
      p_immi_index < 30 ~ "Liberal",
      p_immi_index >= 30 & p_immi_index < 70 ~ "Neutral",
      p_immi_index >= 70 ~ "Restrictive"
    ),
    
    # Trust in politicians (0-10 scale)
    trust_cat = case_when(
      is.na(trstplt) ~ NA_character_,
      trstplt <= 3 ~ "Low trust",
      trstplt >= 4 & trstplt <= 6 ~ "Medium trust",
      trstplt >= 7 ~ "High trust"
    ),
    
    # Left-Right scale (0-10 scale)
    lr_cat = case_when(
      is.na(lrscale) ~ NA_character_,
      lrscale <= 3 ~ "Left",
      lrscale >= 4 & lrscale <= 6 ~ "Center",
      lrscale >= 7 ~ "Right"
    ),
    
    # Household income feelings (1-4 scale)
    income_cat = case_when(
      is.na(hincfel) ~ NA_character_,
      hincfel == 1 ~ "Comfortable",
      hincfel == 2 ~ "Coping",
      hincfel == 3 ~ "Difficult",
      hincfel == 4 ~ "Very difficult"
    ),
    
    # Simplify education
    education_simple = case_when(
      educ_level %in% c("Less than lower secondary", "Lower secondary") ~ "Low",
      educ_level %in% c("Lower tier upper secondary", "Upper tier upper secondary", 
                        "Advanced vocational, sub-degree") ~ "Medium",
      educ_level %in% c("Lower tertiary education, BA level", 
                        "Higher tertiary education, >= MA level") ~ "High"
    ),
    
    # Simplify class
    class_simple = case_when(
      oesch8 %in% c("Production workers", "Service workers") ~ "Working class",
      oesch8 %in% c("Clerks", "Small business owners") ~ "Lower middle",
      oesch8 %in% c("Technical (semi-)professionals", "(Associate) managers") ~ "Upper middle",
      oesch8 %in% c("Socio-cultural (semi-)professionals", 
                    "Self-employed professionals and large employers") ~ "Professional/Managers"
    )
  ) %>%
  # Convert far_right to numeric if it's not already
  mutate(far_right_num = as.numeric(far_right == "Far-right"))

# Calculate proportions for each category
calc_props <- function(data, var) {
  data %>%
    filter(!is.na(.data[[var]])) %>%  # Remove NAs for this variable
    group_by(across(all_of(var))) %>%
    summarise(
      prop = mean(far_right_num, na.rm = TRUE) * 100,
      n = n(),
      se = sd(far_right_num, na.rm = TRUE) / sqrt(n()) * 100,
      .groups = "drop"
    ) %>%
    rename(category = 1)
}

# Get proportions for each demographic
gender_props <- calc_props(demo_comparison, "gender") %>% mutate(variable = "Gender")
age_props <- calc_props(demo_comparison, "age_cat") %>% mutate(variable = "Age")
immi_props <- calc_props(demo_comparison, "immi_cat") %>% mutate(variable = "Immigration attitudes")
trust_props <- calc_props(demo_comparison, "trust_cat") %>% mutate(variable = "Political trust")
lr_props <- calc_props(demo_comparison, "lr_cat") %>% mutate(variable = "Left-Right position")
income_props <- calc_props(demo_comparison, "income_cat") %>% mutate(variable = "Economic situation")
edu_props <- calc_props(demo_comparison, "education_simple") %>% mutate(variable = "Education")
urban_props <- calc_props(demo_comparison, "urban_rural") %>% mutate(variable = "Location")
class_props <- calc_props(demo_comparison, "class_simple") %>% mutate(variable = "Class")

# Combine all
all_props <- bind_rows(
  gender_props, age_props, immi_props, trust_props, lr_props, income_props,
  edu_props, urban_props, class_props
) %>%
  filter(!is.na(category))

# Set reference categories and calculate differences
reference_levels <- tribble(
  ~variable, ~reference,
  "Gender", "Female",
  "Age", "18-34",
  "Immigration attitudes", "Liberal",
  "Political trust", "High trust",
  "Left-Right position", "Center",
  "Economic situation", "Comfortable",
  "Education", "High",
  "Location", "Urban",
  "Class", "Professional/Managers"
)

# Get reference values
reference_values <- all_props %>%
  inner_join(reference_levels, by = "variable") %>%
  filter(category == reference) %>%
  select(variable, ref_prop = prop)

# Calculate differences from reference
plot_data <- all_props %>%
  left_join(reference_values, by = "variable") %>%
  mutate(
    diff = prop - ref_prop,
    ci_lower = diff - 1.96 * se,
    ci_upper = diff + 1.96 * se,
    is_reference = category %in% reference_levels$reference
  ) %>%
  # Order categories within each variable
  mutate(
    category = factor(category, levels = c(
      # Gender
      "Male", "Female",
      # Age
      "70+", "55-69", "35-54", "18-34",
      # Immigration
      "Restrictive", "Neutral", "Liberal",
      # Trust
      "Low trust", "Medium trust", "High trust",
      # Left-Right
      "Right", "Center", "Left",
      # Economic situation
      "Very difficult", "Difficult", "Coping", "Comfortable",
      # Education
      "Low", "Medium", "High",
      # Location
      "Rural", "Urban",
      # Class
      "Working class", "Lower middle", "Upper middle", "Professional/Managers"
    )),
    variable = factor(variable, levels = c(
      "Gender", "Age", "Immigration attitudes", "Political trust", 
      "Left-Right position", "Economic situation", "Education", "Location", "Class"
    ))
  ) %>%
  arrange(variable, category)

# Create the plot
ggplot(plot_data, aes(x = diff, y = category)) +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray40", linewidth = 0.8) +
  geom_segment(aes(x = 0, xend = diff, y = category, yend = category, 
                   color = diff > 0, alpha = !is_reference), 
               linewidth = 4) +
  geom_errorbar(aes(xmin = ci_lower, xmax = ci_upper, alpha = !is_reference), 
                orientation = "y", width = 0.2, linewidth = 0.8) +
  geom_point(aes(alpha = !is_reference), size = 2.5, color = "black") +
  geom_text(data = plot_data %>% filter(is_reference),
            aes(label = "Reference"), x = 0.5, size = 3, hjust = 0, 
            fontface = "italic", color = "gray50") +
  facet_wrap(~variable, ncol = 1, scales = "free_y", strip.position = "left") +
  scale_color_manual(values = c("TRUE" = "#E74C3C", "FALSE" = "#3498DB"), guide = "none") +
  scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.3), guide = "none") +
  scale_x_continuous(
    breaks = seq(-10, 15, 5),
    labels = function(x) paste0(ifelse(x > 0, "+", ""), x, " pp")
  ) +
  labs(
    title = "Far-Right Support: Differences from Reference Categories",
    subtitle = "Greece, Italy, Portugal, Spain (ESS data)",
    x = "← Less likely to support far-right          More likely to support far-right →",
    y = NULL
  ) +
  theme_solarized(base_size = 14) +
  theme(
    strip.placement = "outside",
    strip.text.y.left = element_text(face = "bold", angle = 0, hjust = 1, size = 13),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.x = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "gray40", size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(margin = margin(t = 10), size = 12)
  )

```

## Key Puzzles from Southern Europe

::: incremental
1.  **The Crisis Timing Puzzle** - When economic crisis creates far-right opportunities, what determines the timing of breakthrough? Can crisis alone drive far-right success, or is something else required?

2.  **The Historical Durability Question** - How long can historical legacies (regime type, founding narratives) delay or block far-right emergence? Is there a breaking point?

3.  **The Supply-Demand Relationship** - Across all cases, demand for far-right politics existed before parties mobilized it. What explains when credible supply emerges to meet demand?

4.  **The Mainstream Strategy Question** - Do different mainstream responses (isolation, incorporation, accommodation) permanently block far-right, or do they merely delay breakthrough?

5.  **The Path Dependency Puzzle** - Does the *type* of regime transition (left-wing revolution vs. right-wing elite transition vs. external collapse) shape far-right trajectories for decades? If so, why?

6.  **The Convergence Question** - Why do regions with dramatically different historical contexts eventually converge toward similar far-right electoral levels (15-25%)?

7.  **The Legitimization Mechanism** - What transforms far-right parties from pariah to mainstream? Does it require government participation, gradual normalization, or mainstream accommodation?
:::

## Questions?
