---
title: "Far-Right in Western Europe"
subtitle: "The Far Right (Week 2)"
author: "Dr. Christos Vrakopoulos"
format:
  revealjs:
    theme: solarized
    slide-number: true
    incremental: true
    
    transition: fade
    fontsize: 24px
    css: custom.css
editor: visual
---

```{r setup, include=FALSE}
# Load required packages
library(dplyr)
library(ggplot2)
library(plotly)
library(tidyverse)
library(manifestoR)
library(lubridate)
library(ggthemes)

# Set Manifesto Project API key
mp_setapikey(key = "193c54d64b1189c526214074350ead07")

# Load Manifesto Project data
mp_data <- manifestoR::mp_maindataset()

# Define party codes for far-right parties
party_codes <- tribble(
  ~party, ~party_label,
  31720, "FN/RN",
  41953, "AfD",
  42420, "FPÖ",
  42710, "BZÖ",
  22710, "CP",
  22711, "CD",
  22720, "LPF",
  22722, "PVV",
  22730, "FvD",
  21914, "VB",
  21917, "VB"
)

# Extract far-right party data
far_right_raw <- mp_data %>%
  filter(party %in% party_codes$party) %>%
  mutate(year = year(as.Date(edate))) %>%
  left_join(party_codes, by = "party") %>%
  select(countryname, year, edate, party_label, pervote, absseat) %>%
  filter(!is.na(pervote)) %>%
  arrange(countryname, year, edate)

# Combine vote shares within same election
far_right_combined <- far_right_raw %>%
  group_by(countryname, year, edate) %>%
  summarise(
    vote_share = sum(pervote, na.rm = TRUE),
    parties = paste(unique(party_label), collapse = " + "),
    .groups = "drop"
  ) %>%
  group_by(countryname, year) %>%
  slice_max(edate, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(country = countryname, year, vote_share)

far_right_data <- far_right_combined

# Individual country datasets
france_data <- far_right_raw %>% 
  filter(countryname == "France") %>%
  group_by(countryname, year, edate) %>%
  summarise(vote_share = sum(pervote, na.rm = TRUE), .groups = "drop") %>%
  group_by(countryname, year) %>%
  slice_max(edate, n = 1) %>%
  ungroup() %>%
  select(country = countryname, year, vote_share)

austria_data <- far_right_raw %>% 
  filter(countryname == "Austria") %>%
  group_by(countryname, year, edate) %>%
  summarise(vote_share = sum(pervote, na.rm = TRUE), .groups = "drop") %>%
  group_by(countryname, year) %>%
  slice_max(edate, n = 1) %>%
  ungroup() %>%
  select(country = countryname, year, vote_share)

germany_data <- far_right_raw %>% 
  filter(countryname == "Germany") %>%
  group_by(year, edate) %>%
  summarise(vote_share = sum(pervote, na.rm = TRUE), .groups = "drop") %>%
  group_by(year) %>%
  slice_max(edate, n = 1) %>%
  ungroup() %>%
  mutate(country = "Germany") %>%
  select(country, year, vote_share)

netherlands_data <- far_right_raw %>% 
  filter(countryname == "Netherlands") %>%
  group_by(year, edate) %>%
  summarise(vote_share = sum(pervote, na.rm = TRUE), .groups = "drop") %>%
  group_by(year) %>%
  slice_max(edate, n = 1) %>%
  ungroup() %>%
  mutate(country = "Netherlands") %>%
  select(country, year, vote_share)

belgium_data <- far_right_raw %>% 
  filter(countryname == "Belgium") %>%
  group_by(year, edate) %>%
  summarise(vote_share = sum(pervote, na.rm = TRUE), .groups = "drop") %>%
  group_by(year) %>%
  slice_max(edate, n = 1) %>%
  ungroup() %>%
  mutate(country = "Belgium") %>%
  select(country, year, vote_share)
```

## Important Announcements {vertical-align="center"}

::: {.callout-important appearance="default" icon="false"}
## Session Cancellation

The **Week 5 session** is cancelled.
:::

<br>

::: {.callout-note appearance="simple"}
## Course Schedule

Please check the most up-to-date course schedule on **Learn**.
:::

## Today's Lecture

::: incremental
-   **Part 1:** The Puzzle
-   **Part 2:** Five Trajectories
-   **Part 3:** Comparative Synthesis
:::

# Part 1: The Puzzle

## Why Western Europe?

::::::::: columns
::::: {.column width="50%"}
::: {.fragment .fade-in}
#### First region where modern far-right broke through
:::

::: incremental
-   Established, wealthy democracies
-   France, Germany, Austria, Netherlands, Belgium
-   1980s-1990s: similar conditions
:::
:::::

::::: {.column width="50%"}
::: {.fragment .fade-in}
#### Yet timing varies dramatically:
:::

::: incremental
-   France: 1984
-   Austria: 1986
-   Belgium: 1991
-   Netherlands: 2002
-   Germany: 2013/2017
:::
:::::
:::::::::

-   *Given broadly similar exposure to deindustrialisation, immigration, and European integration, why do radical-right parties become electorally relevant at such different times?*


## The academic attention

```{r}
library(rvest)
library(tidyverse)
library(stringr)
library(tidytext)
library(wordcloud)
library(RColorBrewer)

# ============================================================================
# Text Mining: Extreme Right Bibliography - Papers by Year and Country
# ============================================================================

# URL to the bibliography
url <- "https://www.kai-arzheimer.com/extreme-right-western-europe-bibliography/"

# Fetch and parse the HTML
page <- read_html(url)

# Get the main content div containing the bibliography
# The bibliography appears to be in the main content area
full_text <- page %>%
  html_text2()

# Remove the intro/header text (everything before the first author name entry)
# Look for the start of actual bibliography entries (typically start with capital letter, then comma)
text_start <- str_locate(full_text, "Aarøe, Lene")[1, 1]

if (is.na(text_start)) {
  # Fallback: find the first reference pattern
  text_start <- 1
}

bibliography_text <- substr(full_text, text_start, nchar(full_text))

# Split by [BibTeX] which marks the end of each reference entry
references <- bibliography_text %>%
  str_split("\\[BibTeX\\]") %>%
  unlist() %>%
  str_trim()

# Keep only references that look like actual entries (contain a year in parentheses)
references <- references[str_detect(references, "\\(\\d{4}\\)")]

# Create dataframe
ref_df <- tibble(reference = references)

# Extract year - looks for (YYYY) pattern
ref_df <- ref_df %>%
  mutate(
    year = str_extract(reference, "\\((\\d{4})\\)", group = 1),
    year = as.numeric(year)
  ) %>%
  # Define the countries/regions of interest with expanded keywords
  mutate(
    western_europe = str_detect(reference, regex("Western Europe", ignore_case = TRUE)),
    germany = str_detect(reference, regex("\\b(Germany|German|AfD|Alternative|Pegida|Identitaer|Identitarian)\\b", ignore_case = TRUE)),
    austria = str_detect(reference, regex("\\b(Austria|Austrian|FPÖ|Haider|Hofer)\\b", ignore_case = TRUE)),
    belgium = str_detect(reference, regex("\\b(Belgium|Belgian|Vlaams Belang|Vlaams|Antwerpen|Flanders)\\b", ignore_case = TRUE)),
    france = str_detect(reference, regex("\\b(France|French|Front National|Marine Le Pen|Le Pen|Rassemblement National|RN)\\b", ignore_case = TRUE)),
    netherlands = str_detect(reference, regex("\\b(Netherlands|Dutch|PVV|Geert Wilders|Wilders)\\b", ignore_case = TRUE))
  ) %>%
  # Create a flag for any mention of target regions/countries
  mutate(
    target_region = western_europe | germany | austria | belgium | france | netherlands
  ) %>%
  # Filter to only those with target regions
  filter(target_region == TRUE & !is.na(year))

# Summary: Count papers by year
papers_by_year <- ref_df %>%
  group_by(year) %>%
  summarise(
    count = n(),
    .groups = "drop"
  ) %>%
  arrange(year)

ggplot(papers_by_year, aes(x = year, y = count)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_line(color = "darkred", linewidth = 1) +
  geom_point(color = "darkred", size = 3) +
  labs(
    title = "Papers on Extreme Right in Western Europe by Year",
    subtitle = "References to: Western Europe, Germany, Austria, Belgium, France, Netherlands",
    x = "Year",
    y = "Number of Papers",
    caption = "Source: Kai Arzheimer Bibliography"
  ) +
  theme_solarized() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


```

# Part 2: Five Trajectories

## France: Front National / Rassemblement National {background-image="images/marine_jean.jpeg" background-size="cover" background-opacity="0.3"}

::::: {.column width="50%"}
::: {.fragment .fade-in}
#### Origins:
:::

-   Founded 1972 (Jean-Marie Le Pen)

-   United nationalists, Poujadists, ex-OAS (Algeria)

-   Fringe until 1983 (\<1%)

-   Holocaust denial

::: {.fragment .fade-in}
#### Breakthrough:
:::

-   1983 Dreux: Centre-right coalition with FN

-   1984: 11% European Parliament
:::::

:::: {.column width="50%"}
::: {.fragment .fade-in}
#### Core Ideology:
:::

-   **Leadership:** Jean-Marie (1972-2011) → Marine (2011-present)

-   **Nativism:** Anti-immigration (esp. North African)

-   **Economic:** Left turn under Marine (welfare chauvinism)

-   **Social:** Traditional/Catholic → Secular under Marine

-   **EU:** Hard Eurosceptic → "Reform from within"
::::



## France: Trajectory & Pattern

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

p_france <- ggplot(france_data, aes(x = year, y = vote_share)) +
  geom_line(size = 1.5, color = "#E74C3C") +
  geom_point(size = 3, color = "#E74C3C") +
  geom_vline(xintercept = c(1984, 2011), linetype = "dashed", alpha = 0.5) +
  annotate("text", x = 1986, y = 12, label = "Breakthrough (1984)", hjust = -0.1, size = 4) +
  annotate("text", x = 2011, y = 17, label = "Marine", hjust = -0.1, size = 4) +
  annotate("text", x = 2022, y = 23, label = "41.5%\nrunoff", hjust = -0.1, size = 4) +
  labs(x = "Year", y = "Vote Share (%)") +
  theme_solarized(base_size = 16)

ggplotly(p_france, tooltip = c("x", "y")) %>%
  layout(autosize = FALSE, width = 900, height = 500)
```



## Austria: Freedom Party (FPÖ) & BZÖ {background-image="images/haider_fpo.jpg" background-size="cover" background-opacity="0.2"}

:::::::: columns
::::: {.column width="50%"}
::: {.fragment .fade-in}
#### Origins:
:::

-   Founded 1956 (former Nazis, pan-German nationalists)

-   Third "camp" in Austrian politics (alongside Social Democrats and Christian Democrats)

-   Member of Liberal International (!!) → shows pre-Haider ideological ambiguity - Stable 5% (1956-1986)

::: {.fragment .fade-in}
#### 1986: Jörg Haider revolution
:::

-   Age 36, charismatic, provocative

-   Shifts right: anti-immigration, anti-EU
:::::

:::: {.column width="50%"}
::: {.fragment .fade-in}
#### Core Ideology:
:::

-   **Nativism:** Anti-immigration (Muslims, Eastern Europeans)

-   **Economic:** Traditionally -\> Welfare chauvinism (Rathgeb 2021: ‘exclusive solidarity’)

-   **Social:** Traditional Catholic

-   **EU:** Soft Eurosceptic: critical of integration, but not outright exit

-   **Key leaders:** Haider (1986-2005) → Split → Strache (2005-2019) → Kickl (2021-)
::::
::::::::


## Austria: Trajectory & Pattern

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

p_austria <- ggplot(austria_data, aes(x = year, y = vote_share)) +
  geom_line(size = 1.5, color = "#E67E22") +
  geom_point(size = 3, color = "#E67E22") +
  annotate("rect", xmin = 2000, xmax = 2005, ymin = 0, ymax = 30, alpha = 0.15, fill = "red") +
  annotate("rect", xmin = 2017, xmax = 2019, ymin = 0, ymax = 30, alpha = 0.15, fill = "red") +
  annotate("text", x = 2002.5, y = 28, label = "In govt", size = 3.5, color = "red") +
  annotate("text", x = 2018, y = 28, label = "In govt", size = 3.5, color = "red") +
  geom_vline(xintercept = c(1986, 2005), linetype = "dashed", alpha = 0.5) +
  labs(x = "Year", y = "Vote Share (%)") +
  theme_solarized(base_size = 16)

ggplotly(p_austria, tooltip = c("x", "y")) %>%
  layout(autosize = FALSE, width = 900, height = 500)
```


## Netherlands: Succession of Parties {background-image="images/wilders.jpeg" background-size="cover" background-opacity="0.2"}

:::::::: columns
::::: {.column width="50%"}
::: {.fragment .fade-in}
#### Failed extreme parties (1980-2002):
:::

-   Centre Party (CP): Neo-Nazi, banned 1998

-   Centre Democrats (CD): Radical-right; never more than 3 seats

-   Marginalised as too racist for Dutch voters and competitors

::: {.fragment .fade-in}
#### 2002: Pim Fortuyn changes game
:::

-   Personally liberal on drugs, sexuality, gay rights (Fortuyn was openly gay)

-   Simultaneously: sharply anti-Islam, anti-multiculturalism

-   Assassinated 9 days before election

-   LPF wins 17% but then rapidly collapses (internal chaos, leader loss)
:::::

:::: {.column width="50%"}
::: {.fragment .fade-in}
#### PVV - Wilders:
:::

-   Founded 2006 as Geert Wilders’ personal vehicle (no formal members).

-   **Nativism:** Anti‑Islam, anti‑Muslim & some other migrants

-   **Economic:** Welfare‑chauvinist, blurry but not clearly neoliberal

-   **Social:** Liberal rhetoric, authoritarian in practice

-   **EU:** Hard, identity‑driven Euroscepticism
::::
::::::::

-   **Key insights**: far-right fragmented due to electoral system; new players entered the game Forum for Democracy (FvD), JA21, BBB


## Netherlands: Trajectory & Pattern

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

netherlands_annotated <- netherlands_data %>%
  mutate(
    period = case_when(
      year <= 1998 ~ "CP/CD",
      year > 1998 & year <= 2006 ~ "LPF",
      year > 2006 ~ "PVV/FvD"
    )
  )

p_netherlands <- ggplot(netherlands_annotated, aes(x = year, y = vote_share)) +
  geom_line(size = 1.5, color = "#3498DB") +
  geom_point(aes(shape = period), size = 3, color = "#3498DB") +
  scale_shape_manual(values = c(1, 2, 16), name = "Period") +
  annotate("text", x = 2002, y = 18, label = "Fortuyn", hjust = -0.1, size = 4) +
  annotate("text", x = 2023, y = 24, label = "PVV wins", hjust = 1.1, size = 4) +
  labs(x = "Year", y = "Vote Share (%)") +
  theme_solarized(base_size = 16) +
  theme(legend.position = "bottom")

ggplotly(p_netherlands, tooltip = c("x", "y")) %>%
  layout(autosize = FALSE, width = 900, height = 500)
```


## Germany: Alternative for Germany (AfD) {background-image="images/afd.jpg" background-size="cover" background-opacity="0.1"}

:::::::::: columns
:::::: {.column width="50%"}
::: {.fragment .fade-in}
#### The 30-year puzzle:
:::

-   NPD/Republicans: neo‑Nazi / extreme right, never federal → stigma + cordon sanitaire

-   2013 AfD: soft Euro‑sceptic, market‑liberal, conservative, “respectable” professors

-   2015: refugee crisis + leadership split → radicalization, immigration central

::: {.fragment .fade-in}
#### 2013: AfD founded (NOT far-right initially)
:::

-   Soft Eurosceptic (anti-Euro bailout)

-   Economists, professors

-   Respectable image

::: {.fragment .fade-in}
#### 2015: TRANSFORMATION
:::

-   Internal conflict between Lucke’s moderate, economics‑driven wing and a national‑populist wing culminates in the 2015 split

-   Refugee crisis (1 million arrivals)
::::::

::::: {.column width="50%"}
::: {.fragment .fade-in}
#### Core Ideology (post-2015):
:::

-   **Nativism:** Anti‑immigration, anti‑Islam, refugee “threat”

-   **Economic:** Mixed: market‑liberal + welfare chauvinism

-   **Social:** Conservative/authoritarian, “liberal” rhetoric vs Islam

-   **EU:** From soft to harder Eurosceptic, system‑critical

::: {.fragment .fade-in}
#### Why 30-year delay?
:::

-   Past + stigma + cordon sanitaire delayed a PRRP despite similar demand conditions

-   A moderate, Euro‑issue entry + a salient crisis (refugees 2015/16) were needed to “normalize” the far right and crystallize a new cleavage
:::::
::::::::::

## Germany: Trajectory & Pattern

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

p_germany <- ggplot(germany_data, aes(x = year, y = vote_share)) +
  geom_line(size = 1.5, color = "#34495E") +
  geom_point(size = 3, color = "#34495E") +
  geom_hline(yintercept = 5, linetype = "dashed", color = "red", alpha = 0.5) +
  annotate("text", x = 2010, y = 5.5, label = "5% threshold", size = 3, color = "red") +
  annotate("text", x = 2017, y = 14, label = "Breakthrough", hjust = -0.1, size = 4, fontface = "bold") +
  annotate("text", x = 2015, y = 2, label = "Refugee\ncrisis", size = 3.5, color = "darkred") +
  labs(x = "Year", y = "Vote Share (%)") +
  theme_solarized(base_size = 16)

ggplotly(p_germany, tooltip = c("x", "y")) %>%
  layout(autosize = FALSE, width = 900, height = 500)
```


## Belgium: Flemish Bloc/Interest (VB) {background-image="images/vb.jpg" background-size="cover" background-opacity="0.1"}

::::::::: columns
::::: {.column width="50%"}
::: {.fragment .fade-in}
#### Unique context:
:::

-   Flanders (Dutch, 60%) vs. Wallonia (French, 40%)

-   Separate party systems

-   VB = Flemish-only party

::: {.fragment .fade-in}
#### Origins (Vlaams Blok):
:::

-   1978: Split from VU (moderate Flemish nationalists)

-   Radical wing: independence NOW

-   1991: "Black Sunday" breakthrough (6.6%)

-   Banned (2004), immediately rebranded as "Flemish Interest"
:::::

::::: {.column width="50%"}
::: {.fragment .fade-in}
#### 1989-present: Cordon sanitaire
:::

-   All parties sign agreement

-   NEVER form coalition with VB

-   It limits power but also reinforces VB’s anti‑establishment populism

::: {.fragment .fade-in}
#### Core Ideology VB:
:::

-   **Nativism:** Anti‑immigration + anti‑Walloons

-   **Nationalism:** Flemish separatist

-   **Economic:** Right‑wing, economic nativism

-   **Social:** Authoritarian‑conservative

-   **EU:** Radical‑right Eurosceptic
:::::
:::::::::


## Belgium: Trajectory & Pattern

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

p_belgium <- ggplot(belgium_data, aes(x = year, y = vote_share)) +
  geom_line(size = 1.5, color = "#F39C12") +
  geom_point(size = 3, color = "#F39C12") +
  annotate("rect", xmin = 1989, xmax = 2024, ymin = 0, ymax = 25, alpha = 0.1, fill = "red") +
  annotate("text", x = 2007, y = 23, label = "Cordon sanitaire", size = 3.5, color = "red") +
  geom_vline(xintercept = 2004, linetype = "dashed", color = "black") +
  annotate("text", x = 2004, y = 25, label = "Ban→Rebrand", hjust = -0.1, size = 3.5) +
  labs(x = "Year", y = "Vote Share (%)") +
  theme_solarized(base_size = 16)

ggplotly(p_belgium, tooltip = c("x", "y")) %>%
  layout(autosize = FALSE, width = 900, height = 500)
```


# Part 3: Comparative Synthesis

## Overall Patterns

```{r}
#| echo: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

far_right_flags <- far_right_data %>% 
  filter(country %in% c("France", "Germany", "Austria", "Netherlands", "Belgium")) %>%
  mutate(
    flag_emoji = case_when(
      country == "France" ~ "🇫🇷",
      country == "Germany" ~ "🇩🇪",
      country == "Austria" ~ "🇦🇹",
      country == "Netherlands" ~ "🇳🇱",
      country == "Belgium" ~ "🇧🇪"
    )
  )

# Get only the last year for each country
last_points <- far_right_flags %>%
  group_by(country) %>%
  slice_max(year, n = 1) %>%
  ungroup()

p <- ggplot(far_right_flags, 
            aes(x = year, y = vote_share, color = country, group = country)) +
  geom_line(size = 1.5, alpha = 0.8) +
  geom_point(size = 3, alpha = 0.7) +  # Optional: keep dots for all points
  geom_text(data = last_points,  # ONLY last points get flags
            aes(label = flag_emoji), 
            size = 8, 
            hjust = -0.5) +  # Offset slightly to the right
  scale_color_manual(
    values = c("France" = "#E74C3C", "Germany" = "#34495E", 
               "Austria" = "#E67E22", "Netherlands" = "#3498DB",
               "Belgium" = "#F39C12")
  ) +
  labs(x = "Year", y = "Vote Share (%)") +
  theme_solarized(base_size = 16) +
  theme(legend.position = "bottom")

ggplotly(p) %>%
  layout(autosize = FALSE, width = 900, height = 500)
```



## Timing Variation: Breakthrough vs. Current Power {vertical-align="center"}

::: {style="font-size: 0.85em;"}
| Country | Breakthrough | Context | Status Today (2024-2026) |
|:-----------------|:----------------:|:-----------------|:-----------------|
| **France** | 1984 | Dreux coalition | **RN is the leading opposition force.** Jordan Bardella dominates 2027 presidential polls (\~35%) amid Le Pen's legal appeals. |
| **Austria** | 1986 | Haider leadership | **2024 Election Winner (28.8%).** Currently in opposition after being excluded by the ÖVP-SPÖ-NEOS coalition. |
| **Belgium** | 1991 | "Black Sunday" | **Second largest in Flanders.** Vlaams Belang remains isolated by the *cordon sanitaire* despite high regional polling. |
| **Netherlands** | 2002 | Fortuyn murder | **Largest party in Parliament.** Left the Schoof cabinet (3rd June 2025), triggering a major political crisis. |
| **Germany** | 2017 | Refugee crisis | **Bundestag's second largest party.** Secured 20.8% in the 2025 snap election; currently the primary opposition to the Merz (CDU) gov. |

: {tbl-colwidths="\[15, 15, 20, 50\]"}
:::

. . .

::: {.fragment .fade-up style="background-color: #f8f9fa; padding: 15px; border-left: 5px solid #2c3e50; margin-top: 20px; font-size: 0.8em;"}
**Key Synthesis:** While grievances (demand) remain constant across these wealthy democracies, the **timing of the breakthrough** depends on the **supply**: credible leaders, organizational shifts, and mainstream party reactions.
:::



## Who Votes Far-Right? (ESS Data)

```{r}
#| echo: false
#| warning: false
#| fig-width: 12
#| fig-height: 8

load("/Users/cvrakopo/My Drive/Publications/ESS Data 1-10/Final Creation/ess_analysis_full.RData")  # Adjust path as needed

# Filter data for the five countries
ess_five <- ess_analysis %>%
  filter(country %in% c("Germany", "Austria", "Belgium", "Netherlands", "France"))

# Prepare data with age categories and calculate differences from reference
demo_comparison <- ess_five %>%
  mutate(
    # Create age categories
    age_cat = case_when(
      agea < 35 ~ "18-34",
      agea >= 35 & agea < 55 ~ "35-54",
      agea >= 55 & agea < 70 ~ "55-69",
      agea >= 70 ~ "70+"
    ),
    
    # Immigration attitudes
    immi_cat = case_when(
      is.na(p_immi_index) ~ NA_character_,
      p_immi_index < 30 ~ "Liberal",
      p_immi_index >= 30 & p_immi_index < 70 ~ "Neutral",
      p_immi_index >= 70 ~ "Restrictive"
    ),
    
    # Trust in politicians (0-10 scale)
    trust_cat = case_when(
      is.na(trstplt) ~ NA_character_,
      trstplt <= 3 ~ "Low trust",
      trstplt >= 4 & trstplt <= 6 ~ "Medium trust",
      trstplt >= 7 ~ "High trust"
    ),
    
    # Left-Right scale (0-10 scale)
    lr_cat = case_when(
      is.na(lrscale) ~ NA_character_,
      lrscale <= 3 ~ "Left",
      lrscale >= 4 & lrscale <= 6 ~ "Center",
      lrscale >= 7 ~ "Right"
    ),
    
    # Household income feelings (1-4 scale)
    income_cat = case_when(
      is.na(hincfel) ~ NA_character_,
      hincfel == 1 ~ "Comfortable",
      hincfel == 2 ~ "Coping",
      hincfel == 3 ~ "Difficult",
      hincfel == 4 ~ "Very difficult"
    ),
    
    # Simplify education
    education_simple = case_when(
      educ_level %in% c("Less than lower secondary", "Lower secondary") ~ "Low",
      educ_level %in% c("Lower tier upper secondary", "Upper tier upper secondary", 
                        "Advanced vocational, sub-degree") ~ "Medium",
      educ_level %in% c("Lower tertiary education, BA level", 
                        "Higher tertiary education, >= MA level") ~ "High"
    ),
    
    # Simplify class
    class_simple = case_when(
      oesch8 %in% c("Production workers", "Service workers") ~ "Working class",
      oesch8 %in% c("Clerks", "Small business owners") ~ "Lower middle",
      oesch8 %in% c("Technical (semi-)professionals", "(Associate) managers") ~ "Upper middle",
      oesch8 %in% c("Socio-cultural (semi-)professionals", 
                    "Self-employed professionals and large employers") ~ "Professional/Managers"
    )
  ) %>%
  # Convert far_right to numeric if it's not already
  mutate(far_right_num = as.numeric(far_right == "Far-right"))

# Calculate proportions for each category
calc_props <- function(data, var) {
  data %>%
    filter(!is.na(.data[[var]])) %>%  # Remove NAs for this variable
    group_by(across(all_of(var))) %>%
    summarise(
      prop = mean(far_right_num, na.rm = TRUE) * 100,
      n = n(),
      se = sd(far_right_num, na.rm = TRUE) / sqrt(n()) * 100,
      .groups = "drop"
    ) %>%
    rename(category = 1)
}

# Get proportions for each demographic
gender_props <- calc_props(demo_comparison, "gender") %>% mutate(variable = "Gender")
age_props <- calc_props(demo_comparison, "age_cat") %>% mutate(variable = "Age")
immi_props <- calc_props(demo_comparison, "immi_cat") %>% mutate(variable = "Immigration attitudes")
trust_props <- calc_props(demo_comparison, "trust_cat") %>% mutate(variable = "Political trust")
lr_props <- calc_props(demo_comparison, "lr_cat") %>% mutate(variable = "Left-Right position")
income_props <- calc_props(demo_comparison, "income_cat") %>% mutate(variable = "Economic situation")
edu_props <- calc_props(demo_comparison, "education_simple") %>% mutate(variable = "Education")
urban_props <- calc_props(demo_comparison, "urban_rural") %>% mutate(variable = "Location")
class_props <- calc_props(demo_comparison, "class_simple") %>% mutate(variable = "Class")

# Combine all
all_props <- bind_rows(
  gender_props, age_props, immi_props, trust_props, lr_props, income_props,
  edu_props, urban_props, class_props
) %>%
  filter(!is.na(category))

# Set reference categories and calculate differences
reference_levels <- tribble(
  ~variable, ~reference,
  "Gender", "Female",
  "Age", "18-34",
  "Immigration attitudes", "Liberal",
  "Political trust", "High trust",
  "Left-Right position", "Center",
  "Economic situation", "Comfortable",
  "Education", "High",
  "Location", "Urban",
  "Class", "Professional/Managers"
)

# Get reference values
reference_values <- all_props %>%
  inner_join(reference_levels, by = "variable") %>%
  filter(category == reference) %>%
  select(variable, ref_prop = prop)

# Calculate differences from reference
plot_data <- all_props %>%
  left_join(reference_values, by = "variable") %>%
  mutate(
    diff = prop - ref_prop,
    ci_lower = diff - 1.96 * se,
    ci_upper = diff + 1.96 * se,
    is_reference = category %in% reference_levels$reference
  ) %>%
  # Order categories within each variable
  mutate(
    category = factor(category, levels = c(
      # Gender
      "Male", "Female",
      # Age
      "70+", "55-69", "35-54", "18-34",
      # Immigration
      "Restrictive", "Neutral", "Liberal",
      # Trust
      "Low trust", "Medium trust", "High trust",
      # Left-Right
      "Right", "Center", "Left",
      # Economic situation
      "Very difficult", "Difficult", "Coping", "Comfortable",
      # Education
      "Low", "Medium", "High",
      # Location
      "Rural", "Urban",
      # Class
      "Working class", "Lower middle", "Upper middle", "Professional/Managers"
    )),
    variable = factor(variable, levels = c(
      "Gender", "Age", "Immigration attitudes", "Political trust", 
      "Left-Right position", "Economic situation", "Education", "Location", "Class"
    ))
  ) %>%
  arrange(variable, category)

# Create the plot
ggplot(plot_data, aes(x = diff, y = category)) +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray40", linewidth = 0.8) +
  geom_segment(aes(x = 0, xend = diff, y = category, yend = category, 
                   color = diff > 0, alpha = !is_reference), 
               linewidth = 4) +
  geom_errorbar(aes(xmin = ci_lower, xmax = ci_upper, alpha = !is_reference), 
                orientation = "y", width = 0.2, linewidth = 0.8) +
  geom_point(aes(alpha = !is_reference), size = 2.5, color = "black") +
  geom_text(data = plot_data %>% filter(is_reference),
            aes(label = "Reference"), x = 0.5, size = 3, hjust = 0, 
            fontface = "italic", color = "gray50") +
  facet_wrap(~variable, ncol = 1, scales = "free_y", strip.position = "left") +
  scale_color_manual(values = c("TRUE" = "#E74C3C", "FALSE" = "#3498DB"), guide = "none") +
  scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.3), guide = "none") +
  scale_x_continuous(
    breaks = seq(-20, 30, 5),
    labels = function(x) paste0(ifelse(x > 0, "+", ""), x, " pp")
  ) +
  labs(
    title = "Far-Right Support: Differences from Reference Categories",
    subtitle = "Austria, Belgium, France, Germany and Netherlands (ESS data)",
    x = "← Less likely to support far-right          More likely to support far-right →",
    y = NULL
  ) +
  theme_solarized(base_size = 14) +
  theme(
    strip.placement = "outside",
    strip.text.y.left = element_text(face = "bold", angle = 0, hjust = 1, size = 13),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.x = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "gray40", size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(margin = margin(t = 10), size = 12)
  )
```

## The Big Puzzles

::::: columns
::: {.column width="50%"}
1.  **Germany's 30-year delay – strong latent demand but no credible supply?**

-   Why resistant until 2017?

-   Same conditions as France 1980s

2.  **Austria's government curse**

-   Why collapse when entering govt?

-   Twice! (2000, 2017)

3.  **Belgium's cordon sanitaire**

-   Why isolation works for 35 years?

-   Yet doesn't stop electoral growth
:::

::: {.column width="50%"}
4.  **Netherlands' moderation**

-   Why extremists (CP/CD) failed?

-   But "moderate" anti-Islam succeeded?

5.  **Regional variation**

-   VB only Flanders, not Wallonia

-   Why within-country differences?
:::
:::::

## Patterns We Observe in Western Europe {vertical-align="center"}

::::: {layout="[[1,1]]"}
::: {.fragment .fade-up style="border: 1px solid #ddd; padding: 20px; border-radius: 10px;"}
### Demand-Side Patterns

-   **Timing varies dramatically:** A 30-year span exists between breakthroughs (e.g., France 1984 vs. Germany 2017) despite similar socio-economic conditions.
-   **Historical legacies:** National histories, such as the post-WWII stigma in Germany, shape the "possibility space" for far-right emergence.
-   **Mobilization gaps:** While anti-immigrant or anti-elite attitudes are widespread, parties vary significantly in their ability to mobilize them into votes.
:::

::: {.fragment .fade-up style="border: 1px solid #ddd; padding: 20px; border-radius: 10px;"}
### Supply & Strategic Patterns

-   **Diverse trajectories:** Party development varies across the region: stable (France), volatile (Austria), fragmented (Netherlands), delayed (Germany), and isolated (Belgium).
-   **The Moderation Premium:** Political success correlates with programmatic moderation; open extremism is typically linked to failure.
-   **Mainstream responses:** Outcomes are shaped by how established parties react, ranging from *cordon sanitaire* to accommodation or position theft.
:::
:::::

<br>

::: {.fragment .fade-in style="text-align: center; background-color: #f8f9fa; padding: 15px; border-radius: 8px; border-left: 5px solid #2c3e50; font-weight: bold;"}
**Takeaway:** Far-right success is not automatic: it depends on *how* parties and mainstream actors structure similar grievances.
:::

## Course Roadmap

::::: columns
::: {.column width="50%"}
**Cases (Weeks 2-6):**

-   Week 2: ✓ Western Europe
-   Week 3: UK & Brexit
-   Week 4: Scandinavia
-   Week 5: Southern Europe
-   Week 6: Eastern Europe
:::

::: {.column width="50%"}
**Theory (Weeks 7-10):**

-   Week 7: Demand-side
-   Week 8: Supply-side
-   Week 9: Competition & Normalization
-   Week 10: Global patterns
:::
:::::

. . .

**Pedagogical approach:** Accumulate puzzles from cases → Theory provides systematic answers


## Questions?
